Main Analysis for Biracial Grade Retention Project

### 2022-09-08 analysis/analysis_main.Rmd
Correct all models from main analysis, but not sensitivity tests yet

```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. Non-overlap in color-coded confidence bands roughly indicates statistically significant difference at the 5% level"}
results_mod0$coef_table %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
                ymax=upper, fill=colors), color=NA, alpha=0.2)+
  geom_linerange(data=filter(results_mod0$ci_df, mrace=="Black/White"), 
                 position=position_dodge2(width = 0.05), size=1.2,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point(size=2)+
  coord_flip()+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=color_choices[c(2,5:6)])+
  scale_fill_manual(values=color_choices[c(2,5:6)])+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())
```

@@

results_mod0$coef_table %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
                ymax=upper, fill=colors), color=NA, alpha=0.2)+
  geom_linerange(data=results_mod0$ci_df, 
                 position=position_dodge2(width = 0.05), size=1.2,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point(size=2)+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())
```

As the table here shows, after controlling for other stuff, the majority of multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate. The only case where the biracial group is more like the higher monoracial group is for Latino/Asians. The Indigenous/Latino group is the only group statistically disginguishable as midway between the two constituent groups. The White/Indigenous group is also statistically distinguishable from both constitutent groups but much closer to Whites in their probability. The Black/Asian and Black/Indigenous groups also give point estimates at the midpoint, but the results are very noisy due to large SEs and so conclusions here would have to be speculative.

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
coef_table_mod1 <- calculate_marg_means(model1)
coef_table_mod2 <- calculate_marg_means(model2)

coef_table <- coef_table_mod0 %>%
  mutate(model="crude")
coef_table <- coef_table_mod1 %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- coef_table_mod2 %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- coef_table_mod4 %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")

@@

For the first three models, the results are pretty consistent. Black/White biracials are much closer to the lower grade retention rates of whites, although they have rates that are clearly still higher than whites. Controlling for parental resources changes things considerably. First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half). Second, Black/White biracials actually have lower rates than either group. They aren't that different than whites in effect size but the difference is statistically significant.

@@

Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model. In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.

@@

The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students, although the high SE makes the precision of this effect quite weak. Controlling for cultural esources moves this group closer to the midpoint value. This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well. So this gave them and advantage which is removed in that model. The family resources model moves the Black/Asian group slightly more toward the middle but does not change much from the previous model.

@@

This group starts fairly close to the midway point, but controlling for resources clearly moves them into the "lower half" space. Their results never look as good as whites, but look substantially better than Indigenous students, who continue to have the highest rate of grade retention. Its interesting that the resource models move them closer to white given that they still have resources that are quite a bit worse than Whites. Presumably this is because those resources are still substantially better than Indigenous students.

@@

In all of the models, White/Asian students have grade retention rates lower than the rates of Asians - the monoracial group with the lower rates. Controls for family resources reduce this gap a bit because Asian/White students are the most privileged students overall and the confidence interval with Asian students overlaps slightly. So, we can determine that Asian/White students have grade retention rates at least as low as the lowest consistuent group (Asians) and possibly even lower.

@@

|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
|---------|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| Black/White       |       X\*       |               |            |         |            |                |                  |
| White/Asian       |       X\*       |       X       |            |         |            |                |                  |
| White/Latino      |        X        |      X\*      |            |         |            |                |                  |
| Black/Latino      |        X        |      X\*      |     X      |    X    |            |                |                  |
| Indigenous/Asian  |        X        |      X\*      |     X      |         |            |                |                  |
| Indigenous/White  |                 |               |    X\*     |         |            |                |                  |
| Indigenous/Latino |                 |               |     X      |   X\*   |     X      |                |                  |
| Black Asian       |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Black/Indigenous  |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Latino/Asian      |                 |       X       |     X      |    X    |     X      |       X        |       x\*        |

>>>>

```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. Non-overlap in color-coded confidence bands roughly indicates statistically significant difference at the 5% level"}
results_mod0$coef_table %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
                ymax=upper, fill=colors), color=NA, alpha=0.5)+
  geom_linerange(data=filter(results_mod0$ci_df, mrace=="Black/White"), 
                 position=position_dodge2(width = 0.05), size=1.2, alpha=0.5,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point(size=2)+
  coord_flip()+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=color_choices[c(2,5:6)])+
  scale_fill_manual(values=color_choices[c(2,5:6)])+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())
```

@@

results_mod4$coef_table %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
                ymax=upper, fill=colors), color=NA, alpha=0.5)+
  geom_linerange(data=results_mod4$ci_df, 
                 position=position_dodge2(width = 0.25), size=1, alpha=0.5,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point(size=2)+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())
```

As the table here shows, after controlling for other stuff, the majority of multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate. The only case where the biracial group is more like the higher monoracial group is for Latino/Asians, but the point estimate here is not statistically distinguishable from either monoracial group and both monoracial groups have similar risks after controlling for resources. The Indigenous/Latino, Black/Indigenous, and Black/Asian groups all have point estimates roughly at the halfway pont, but the results are also somewhat inconclusive because of the high SE in all three cases. In the case of the Black/Asian and Black/Indigeious gruops, we cannot rule out that they are higher or lower than either monoracial group. In the Indigenous/Latin group, we cannot rule out that they are like the lower group (Latinos) and the higher one is very, very close. So, basically, in all of the cases where we can make a solid determination, the determination is like the lower group.

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
results_mod1 <- calculate_marg_means(model1)
results_mod2 <- calculate_marg_means(model2)

coef_table <- results_mod0$coef_table %>%
  mutate(model="crude")
coef_table <- results_mod1$coef_table %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- results_mod2$coef_table %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- results_mod4$coef_table %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

ci_df <- results_mod0$ci_df %>%
  mutate(model="crude")
ci_df <- results_mod1$ci_df %>%
  mutate(model="fixed effects") %>%
  bind_rows(ci_df)
ci_df <- results_mod2$ci_df %>%
  mutate(model="cultural resources") %>%
  bind_rows(ci_df)
ci_df <- results_mod4$ci_df %>%
  mutate(model="parent resources") %>%
  bind_rows(ci_df) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

ggplot(coef_table, aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
            color=NA, alpha=0.5)+
  geom_linerange(data=ci_df, 
                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")

@@

For the first three models, the results are pretty consistent. Black/White biracials are much closer to the lower grade retention rates of whites, although they have rates that are clearly still higher than whites. Controlling for parental resources changes things considerably. First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half). Second, Black/White biracials actually have lower rates than either group, although the difference from White students is not statistically significant.

@@

Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model. In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models and is not statistically significantly different from Latinos in the cultural resources model.

@@

The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students, although the high SE makes the precision of this effect quite weak. Controlling for cultural resources moves this group closer to the midpoint value. This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well. So this gave them and advantage which is removed in that model. The family resources model moves the Black/Asian group slightly more toward the middle but does not change much from the previous model.

@@

This group starts fairly close to the midway point, but controlling for resources clearly moves them into the "lower half" space. Their results never look as good as whites (although the bars are so close to overlapping in the full model), but look substantially better than Indigenous students, who continue to have the highest rate of grade retention. Its interesting that the resource models move them closer to white given that they still have resources that are quite a bit worse than Whites. Presumably this is because those resources are still substantially better than Indigenous students.

@@

In all of the models, White/Asian students have grade retention rates lower than the rates of Asians - the monoracial group with the lower rates. Controls for family resources reduce this gap to the point where the difference is not statistically distinguishable from Asian students. So, we can determine that Asian/White students have grade retention rates at least as low as the lowest consistuent group (Asians) and possibly even lower.

@@

|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
| Black/White       |       X\*       |       X       |            |         |            |                |                  |
| White/Asian       |       X\*       |       X       |            |         |            |                |                  |
| White/Latino      |        X        |      X\*      |            |         |            |                |                  |
| Black/Latino      |        X        |      X\*      |     X      |    X    |            |                |                  |
| Indigenous/Asian  |        X\*      |       X       |     X      |    X    |            |                |                  |
| Indigenous/White  |                 |       X       |    X\*     |         |            |                |                  |
| Indigenous/Latino |                 |       X       |     X      |   X\*   |     X      |                |                  |
| Black Asian       |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Black/Indigenous  |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Latino/Asian      |                 |       X       |     X      |    X    |     X      |       X        |       x\*        |
